\\LARGE Decision-Optimal Probabilistic Forecast Reconciliation for Expected Shortfall in Hierarchical Time Series

### 2025-08-25 phd_application/phd_propossal.Rmd
finish up motivation

Risk managers and regulators oftenly use quantitative risk measures to quantify and manage potential losses in uncertain environments. Two familiar choices are Value-at-Risk (VaR) and Expected Shortfall (ES). While VaR is conceptually straightforward, it only measures the percentiles of loss distributions and ignores the severity of losses beyond that threshold [@Yamai2005-me]. ES, on the other hand, captures the average loss in the tail beyond the VaR threshold. This makes ES a more reliable metric for assessing and managing "tail risks", and it has been adopted as the preferred regulatory metric since the Basel III framework [@Basel-iii].

These tail risks are, however, not managed in isolation. Consider a bank's trading book, which consists of multiple portfolios across different desks, where each desk manages various asset classes, and then individual securities. This is an example of a hierarchical structure. Risk managers need to assess and forecast the risk quantities at each aggregation level, from the overall trading book down to individual securities Ignoring this naturally-formed hierarchical structure can lead to poor forecasts of uncertainty, resulting in misallocation of capital and suboptimal risk management strategies.

@@

In practice, when forecasts are produced for all series, they typically violate these aggregation constraints observed in the data. Such forecasts are *incoherent*, which often happens when forecasts at lower-level series (e.g. individual securities) do not sum up to the aggregated series (e.g. overall trading book). This can undermine downstream decisions that require internal consistency. Forecast reconciliation, a post-processing step, addresses this by adjusting the so-called base forecasts so the final adjusted set is *coherent* with the aggregation structure. Approaches such as OLS [@Hyndman2011-jv], WLS [@Hyndman2016-ic], and MinT [@Wickramasuriya2019-xq] implement this as linear mappings so that the adjusted forecasts lie in the coherent subspace; see @Athanasopoulos2024-as for a comprehensive review.

An additional incentive for the rapid adoption of reconciliation methods is that they improve forecast accuracy. For example, the International Monetary Fund reports improvements in liquidity forecasts of $3 million a day in a medium-sized economy, generating economic benefits via more efficient monetary policy operations [@El-Gemayel2022-hx]. This advantage, combined with the need for coherent forecasts, motivates extending reconciliation methods to probabilistic forecasts, especially for tail risk measures such as ES.

@@

Existing reconciliation methods for point forecasts has recently been generalised to reconcile probabilistic forecasts with downstream decisions based on these. However, these approaches originally focus on point forecasts and are optimised for scoring functions such as mean squared error (MSE). There is a gap in the literature on how *optimal* this approach when considering alternative quantities of interest, such as quantiles or expected shortfall (ES). This leaves the question: *"How can we construct probabilistic forecast reconciliation that are explicitly optimised for ES (and, where useful, jointly for VaR and ES)?"*

Although @Panagiotelis2023-fs introduced an score optimisation for reconciliation weights using proper scoring rules, energy score and variogram score, for full multivariate predictive distributions, they do not directly address the tail functionals. Other than that, there is a lack of research on optimising reconciliation for specific risk measures such as VaR and ES.

- briefly expand on limitaions of VaR: not coherent risk measure, not subadditive, on top of ignoring tail risk
- focus on ES
- however ES is not elicitable, but is jointly elicitable with the quantile function
- Bring the work of Fissler and Ziegel (2016) and Rockafellar and Uryasev (2000) into the reconciliation framework

>>>>

<!-- 
While VaR is conceptually straightforward, it only measures the percentiles of loss distributions and ignores the severity of losses beyond that threshold [@Yamai2005-me]. ES, on the other hand, captures the average loss in the tail beyond the VaR threshold. This makes ES a more reliable metric for assessing and managing "tail risks", and it has been adopted as the preferred regulatory metric since the Basel III framework [@Basel-iii].
-->

Risk managers and regulators oftenly use quantitative risk measures to quantify and manage potential losses in uncertain environments. Two familiar choices are Value-at-Risk (VaR) and Expected Shortfall (ES; also known as Conditional VaR). These tail risks are, however, not managed in isolation. Consider a bank's trading book, which consists of multiple portfolios across different desks, where each desk manages various asset classes, and then individual securities. This is an example of a hierarchical structure. Risk managers need to assess and forecast the risk quantities at each aggregation level, from the overall trading book down to individual securities. Ignoring this naturally-formed hierarchical structure can lead to poor forecasts of uncertainty, resulting in misallocation of capital and suboptimal risk management strategies.

@@

In practice, when forecasts are produced for all series, they typically violate these aggregation constraints observed in the data. Such forecasts are *incoherent*, which often happens when forecasts at lower-level series (e.g. individual securities) do not sum up to the aggregated series (e.g. overall trading book). This can undermine downstream decisions that require internal consistency. Forecast reconciliation, a post-processing step, addresses this by adjusting the so-called base forecasts so the final adjusted set is *coherent* with the aggregation structure. Approaches such as OLS [@Hyndman2011-jv], WLS [@Hyndman2016-ic], and MinT [@Wickramasuriya2019-xq] are commonly used in practice; see @Athanasopoulos2024-as for a comprehensive review.

An additional incentive for the rapid adoption of reconciliation methods is that they improve forecast accuracy. For example, the International Monetary Fund reports improvements in liquidity forecasts of $3 million a day in a medium-sized economy, generating economic benefits via more efficient monetary policy operations [@El-Gemayel2022-hx]. This context motivates methods that deliver coherent, multi-level forecasts whose uncertainty aligns with the tail-risk metrics managers actually use.

@@

Existing reconciliation methods for point forecasts has recently been generalised to reconcile probabilistic forecasts with downstream decisions based on these. However, these approaches originally focus on point forecasts and are optimised for scoring functions such as mean squared error (MSE). There is a gap in the literature on how *optimal* this approach when considering alternative quantities of interest, such as Value-at-Risk or Expected Shortfall. This leaves the question: *"How can we construct probabilistic forecast reconciliation that are explicitly optimised for tail-risk measures?"*

Although @Panagiotelis2023-fs introduced an score optimisation for reconciliation weights using proper scoring rules (energy score and variogram score) for full multivariate predictive distributions, they do not directly address the tail functionals. Other than that, there is a lack of research on optimising reconciliation for specific risk measures such as VaR and ES.

Both VaR and ES are popular risk measures, but they have different properties. While VaR is conceptually straightforward, it ignores the severity of losses beyond the VaR threshold and is not a coherent risk measure, since it does not satisfy the subadditivity property [@Yamai2005-me]. VaR can mislead risk assessments across different levels of the hierarchy, and may underestimate the true risk exposure under market stress. ES, on the other hand, is a coherent metric that captures the average loss in the tail beyond the VaR threshold. This makes ES a more reliable metric for assessing and managing "tail risks", and it has been adopted as the preferred regulatory metric since the Basel III framework [@Basel-iii].

Even with these advantages, ES has its own challenges. Notably, ES is not elicitable on its own, meaning there is no scoring rule or loss function that can uniquely identify the ES as the optimal optimiser. However, works from @Fissler2015-uu and @Rockafellar2000-el show that this is not impossible when ES is considered jointly with VaR. This joint elicitability opens the door to optimising reconciliation methods specifically for ES. As a result, the biggest objective of this project is to bring these recent theoretical advances into the reconciliation framework, to develop methods that produce coherent probabilistic forecasts optimised for ES.
